\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{array}
\usepackage{url}
\usepackage{hyperref}

\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

% ─────────────────────────────────────────────────────────────────────────────
\begin{document}

\title{DB-Lighthouse: An AI-Augmented PostgreSQL Schema Analysis and\\
Optimization Platform}

\author{
    \IEEEauthorblockN{Chinmay Tikole}
    \IEEEauthorblockA{
        \textit{Department of Information Technology} \\
        \textit{Vidyalankar Institute of Technology} \\
        Mumbai, India \\
        chinmay.tikole@vit.edu.in
    }
    \and
    \IEEEauthorblockN{Aniket Panchal}
    \IEEEauthorblockA{
        \textit{Department of Information Technology} \\
        \textit{Vidyalankar Institute of Technology} \\
        Mumbai, India \\
        aniket.panchal@vit.edu.in
    }
    \and
    \IEEEauthorblockN{Aditya Ram Vemuri}
    \IEEEauthorblockA{
        \textit{Department of Information Technology} \\
        \textit{Vidyalankar Institute of Technology} \\
        Mumbai, India \\
        aditya.vemuri@vit.edu.in
    }
}

\maketitle

% ─────────────────────────────────────────────────────────────────────────────
\begin{abstract}
Modern web applications increasingly rely on relational database management
systems (RDBMS) as their primary persistence layer, yet database schema
health — encompassing query performance, storage efficiency, security posture,
and structural integrity — remains poorly observable without specialized
expertise. This paper presents DB-Lighthouse, a full-stack, AI-augmented
platform that enables developers of all experience levels to connect a live
PostgreSQL database, receive automated schema analysis, visualize table
relationships, and obtain natural-language explanations and actionable
optimization recommendations. A core safety feature is the \emph{shadow
database pattern}: the system clones only the schema of the user's database
into an isolated local PostgreSQL instance, ensuring production data is never
accessed or modified. Integrated analysis modules cover query performance,
index recommendations, anomaly detection, security and personally identifiable
information (PII) scanning, migration validation, schema drift tracking, and
semantic rule enforcement. An embedded large language model (LLM) translates
every finding into plain-English summaries, making advanced database
observability accessible to non-expert users including product managers and
startup founders. Empirical evaluation across several real-world schemas
demonstrates measurable reductions in estimated storage footprint (up to
20--25\%) and significant improvements in query latency through automated
index recommendations.
\end{abstract}

\begin{IEEEkeywords}
database optimization, schema analysis, large language models, PostgreSQL,
shadow database, index recommendation, anomaly detection, PII detection,
schema visualization, developer tooling
\end{IEEEkeywords}

% ─────────────────────────────────────────────────────────────────────────────
\section{Introduction}

The proliferation of cloud-hosted PostgreSQL services — including Amazon RDS,
Google Cloud SQL, Supabase, Railway, and Neon — has democratized access to
production-grade relational databases for startups and individual developers.
However, this accessibility has not been accompanied by a commensurate
democratization of database observability tooling. Traditional database
administration (DBA) tasks such as index tuning, normalization analysis,
anomaly triage, and compliance scanning remain time-intensive and require
deep expertise in SQL internals, query planning, and systems performance.
As a result, many small engineering teams operate databases that accumulate
technical debt silently: bloated schemas, redundant indexes, sub-optimal
data types, and unscanned PII fields.

Existing commercial products in the database observability space (e.g.,
Datadog, New Relic, PgHero) address subsets of these problems but either
target large enterprises with complex on-boarding requirements, impose
significant cost at modest data volumes, or present information in a form
that is meaningful only to experienced DBAs. Natural language interfaces to
databases have been the subject of considerable academic research
\cite{li2023can, pourreza2024din, gao2023texttosql}, yet integration of
such interfaces with live schema analysis and actionable optimization
recommendations within a unified, secure platform has not been demonstrated
at accessible scale.

This paper makes the following contributions:

\begin{itemize}
    \item A \textbf{shadow database architecture} that decouples analysis
    from production, guaranteeing zero risk of data modification.
    \item An integrated \textbf{multi-domain analysis engine} spanning query
    performance, indexing, anomaly detection, governance, security, and
    semantic rule validation in a single platform.
    \item A \textbf{natural-language interface} powered by a locally-hosted
    LLM, making database insights accessible to non-expert stakeholders.
    \item A \textbf{schema visualization} subsystem using force-directed
    graphs that renders table relationships, size heatmaps, and bottleneck
    indicators interactively.
    \item An \textbf{open evaluation} of platform effectiveness across
    diverse real-world PostgreSQL schemas.
\end{itemize}

The remainder of this paper is organized as follows. Section~II surveys
related work. Section~III describes the system architecture and shadow
database pattern. Section~IV details each analysis module. Section~V
presents the frontend design and user experience. Section~VI discusses
the target user personas and deployment contexts. Section~VII evaluates
system performance. Section~VIII concludes with future directions.

% ─────────────────────────────────────────────────────────────────────────────
\section{Related Work}

\subsection{Text-to-SQL and Natural Language Interfaces}

The problem of translating natural language queries into executable SQL has
been studied extensively. Early rule-based systems such as LUNAR
\cite{woods1973progress} and MASQUE \cite{androutsopoulos1995nldb} demonstrated
feasibility but were brittle in the face of schema heterogeneity. The
introduction of large pre-trained language models significantly advanced
the state of the art: T5-based architectures fine-tuned on the Spider
benchmark \cite{yu2018spider} achieved near-human accuracy on single-database
tasks, while models such as GPT-4 \cite{openai2023gpt4} and Llama 2
\cite{touvron2023llama} enabled zero-shot query generation across novel
schemas. Recent work on DIN-SQL \cite{pourreza2024din} decomposes complex
queries into sub-problems, improving accuracy on multi-table joins. Despite
these advances, integration of natural language interfaces with live schema
introspection and safety guardrails for production databases has received
limited attention.

\subsection{Automated Index Recommendation}

Index selection is a classic problem in physical database design. The
AutoAdmin project \cite{chaudhuri1997autoadmin} introduced workload-driven
index recommendation using query cost estimation, an approach subsequently
adopted in Microsoft SQL Server. More recent work applies reinforcement
learning and graph neural networks to the problem \cite{lan2020index,
basu2020weightedge}. PgHero \cite{pghero2022} provides lightweight
heuristic-based index suggestions for PostgreSQL using \texttt{pg\_stat\_user\_indexes}
but lacks natural-language explanations or integration with broader
observability. DB-Lighthouse extends rule-based heuristics with LLM-generated
rationales that are understandable to developers without query planning expertise.

\subsection{Database Anomaly Detection}

Anomaly detection in OLTP systems has been studied in the context of
performance regression \cite{ma2020diagnosing} and security intrusion
detection \cite{sallam2015dbsafe}. Systems such as OtterTune
\cite{van2017ottertune} tune database configuration parameters using
machine learning to improve throughput. These approaches operate at the
workload or configuration layer and do not address schema-level anomalies.
DB-Lighthouse introduces schema-focused anomaly detection that surfaces
structural drift, unusual growth patterns, and constraint violations.

\subsection{Schema Visualization}

Interactive schema visualization tools include SchemaSpy \cite{schemaspy2022}
and DBeaver's ER diagram generator \cite{dbeaver2023}, which render
entity-relationship diagrams from live databases. These tools are read-only
renderers and provide no optimization guidance. React Flow \cite{reactflow2023},
the visualization library used in DB-Lighthouse, enables force-directed
interactive graphs suitable for real-time rendering of evolving schemas.

\subsection{Database Governance and Compliance}

Data governance frameworks such as Apache Atlas \cite{atlas2023} and
Collibra \cite{collibra2023} address enterprise metadata management and
lineage tracking. PII detection at the database layer is supported by tools
such as AWS Macie \cite{macie2023} and Google Cloud DLP \cite{clouddlp2023},
but these are cloud-provider-specific. DB-Lighthouse provides portable,
schema-level PII heuristic detection independent of cloud provider,
integrated directly into the analysis dashboard.

% ─────────────────────────────────────────────────────────────────────────────
\section{System Architecture}

\subsection{Architectural Overview}

DB-Lighthouse follows a three-tier architecture comprising a Next.js
frontend, a FastAPI backend, and a shadow PostgreSQL instance. All
communication between tiers uses RESTful HTTP with JSON payloads. The
frontend is served as a static Next.js application; the backend exposes
a versioned API under the \texttt{/api} prefix; and the shadow database
runs as a Docker-managed PostgreSQL container isolated from external networks.

\bigskip
\noindent\textbf{[DIAGRAM CUE — Figure 1: High-level three-tier architecture
diagram. Show three boxes: (1)~Next.js Frontend on port~3000, (2)~FastAPI
Backend on port~8000, (3)~Shadow PostgreSQL on port~5432. Draw arrows
for: browser $\leftrightarrow$ frontend, frontend $\leftrightarrow$ backend,
backend $\leftrightarrow$ shadow DB, and a dashed arrow from the user's
external PostgreSQL through the backend (schema-only extraction via
pg\_dump). Label the dashed arrow ``schema-only (no data)'' to emphasize
the safety guarantee.]}
\bigskip

\subsection{Shadow Database Pattern}

The shadow database pattern is the foundational safety mechanism of
DB-Lighthouse. When a user supplies a PostgreSQL connection string, the
backend performs the following sequence:

\begin{enumerate}
    \item Verify connectivity to the source database using a read-only
    probe query.
    \item Execute \texttt{pg\_dump --schema-only} against the source to
    extract Data Definition Language (DDL) statements without any row data.
    \item Restore the DDL into a local shadow PostgreSQL instance using
    \texttt{psql}.
    \item Perform all subsequent analysis queries against the shadow
    instance exclusively.
\end{enumerate}

Let $S$ denote the user's source schema and $S'$ the shadow clone. The
transformation $f: S \rightarrow S'$ preserves the complete structural
fidelity of $S$ — including all tables, columns, constraints, indexes,
foreign keys, sequences, and types — while discarding all row-level data.
Formally, for any table $T \in S$ and its shadow counterpart $T' \in S'$:

\begin{equation}
    |T'| = 0 \quad \text{and} \quad \text{schema}(T') = \text{schema}(T)
\end{equation}

where $|\cdot|$ denotes the row count. This guarantee ensures that sensitive
production data is never transmitted to or stored within the platform at
any stage of processing. User credentials are stored in memory only for the
duration of the active session and are never persisted to disk.

\subsection{Backend Service Architecture}

The backend adopts a service-oriented design in which each analysis domain
is encapsulated by a dedicated service module paired with a corresponding
API endpoint module. Table~\ref{tab:services} summarizes the service-to-endpoint
mapping.

\begin{table}[htbp]
\caption{Backend Service and Endpoint Mapping}
\label{tab:services}
\begin{tabular}{>{\raggedright}p{2.8cm} >{\raggedright}p{2.3cm} >{\raggedright\arraybackslash}p{2.3cm}}
\toprule
\textbf{Domain} & \textbf{Service Module} & \textbf{Endpoint Module} \\
\midrule
DB connection \& cloning & db\_service & db\_connection \\
AI Q\&A & ai\_service & analysis \\
Schema introspection & schema\_analysis & analysis \\
Index analysis & index\_analyzer & optimization \\
Optimization & optimization\_service & optimization \\
Anomaly detection & anomaly\_detector & anomaly \\
Incident tracking & incident\_engine & incidents \\
Migration validation & migration\_analyzer & governance \\
Schema drift & drift\_detector & governance \\
Security / PII & (inline) & security \\
Semantic rules & (inline) & semantic\_rules \\
MCP server & mcp\_server & mcp \\
\bottomrule
\end{tabular}
\end{table}

\subsection{AI Safety Firewall}

All natural-language queries that result in SQL generation pass through
a prompt firewall implemented in \texttt{services/prompt\_firewall.py}.
The firewall applies a rule-based pattern matching stage that blocks
queries containing destructive SQL patterns including \texttt{DROP},
\texttt{TRUNCATE}, \texttt{DELETE} without a \texttt{WHERE} clause, and
\texttt{ALTER TABLE ... DROP COLUMN}. Additionally, any \texttt{SELECT}
statement lacking an explicit \texttt{LIMIT} clause is automatically
augmented with \texttt{LIMIT~100} before execution, preventing accidental
full-table scans that could overwhelm the shadow instance for large schemas.
This two-layer defense ensures that neither injected adversarial prompts
nor inadvertent LLM hallucinations can cause destructive operations.

\subsection{Deployment and Infrastructure}

The full platform is packaged as a Docker Compose application with three
services: the Next.js frontend container, the FastAPI backend container,
and the shadow PostgreSQL container. A single \texttt{docker-compose up
--build} command brings the entire stack to operational status, enabling
immediate local deployment without cloud accounts, API keys, or external
dependencies. The AI module uses Jan AI running a quantized Mistral 7B model
(\texttt{mistral-ins-7b-q4}) by default, routed through an OpenAI-compatible
REST interface, with fallback support for Ollama and direct GGUF file loading.

% ─────────────────────────────────────────────────────────────────────────────
\section{Analysis Modules}

\subsection{Schema Introspection and Relationship Mapping}

Upon connecting a database, the schema analysis service introspects the
shadow instance's information schema and \texttt{pg\_catalog} system tables
to extract a comprehensive metadata graph. For each table the service
collects: column names and data types, nullable flags, primary key designations,
foreign key relationships, existing index definitions, estimated row counts
via \texttt{pg\_class.reltuples}, and estimated storage sizes via
\texttt{pg\_relation\_size}. This metadata is serialized as a JSON graph
structure in which nodes represent tables and directed edges represent
foreign key references.

\bigskip
\noindent\textbf{[DIAGRAM CUE — Figure 2: Screenshot of the interactive
schema visualization graph (React Flow canvas) on the Dashboard page.
Capture a real database schema showing several tables with colored nodes
(green = low storage, yellow = medium, red = high) and edges of varying
thickness representing FK relationships. Include the right-side detail
panel showing column breakdown and index list for a selected table.]}
\bigskip

The graph rendering engine combines React Flow for node/edge layout management
with D3-force for physics-based positioning and Dagre for hierarchical
layout. Node size is proportional to estimated row count, and node color
encodes storage utilization on a green–yellow–red heat scale. Edge thickness
is weighted by the count of foreign key references between a pair of tables.
A ``Highlight Bottlenecks'' toggle flags tables with five or more inbound
foreign key references, which are common sources of lock contention in
high-write workloads.

\subsection{Storage Optimization Engine}

The storage optimization engine performs a statistical scan of each table
to identify data type inefficiencies and normalization opportunities. For
each column, the engine computes:

\begin{equation}
    \mu_l = \frac{1}{N}\sum_{i=1}^{N} \text{len}(v_i),\quad
    \sigma_l = \sqrt{\frac{1}{N}\sum_{i=1}^{N}(\text{len}(v_i)-\mu_l)^2}
\end{equation}

where $v_i$ is the $i$-th sampled value and $N \leq 1000$ is the sample
size. Using $\mu_l$ and $\sigma_l$, the engine applies the following
heuristic rules:

\begin{itemize}
    \item A \texttt{TEXT} column with $\mu_l < 255$ is flagged as a
    candidate for conversion to \texttt{VARCHAR(\lceil\mu_l + 3\sigma_l\rceil)}$.
    \item A \texttt{VARCHAR(k)} column where $\mu_l < 0.1k$ is flagged
    for tighter sizing.
    \item A nullable column with a null rate of $0\%$ across the sample
    is flagged for \texttt{NOT NULL} constraint enforcement.
    \item Columns whose distinct value count is $\leq 20$ relative to sample
    size $N$ are candidates for enum normalization.
    \item Unused indexes (zero scans in \texttt{pg\_stat\_user\_indexes})
    are flagged for removal.
    \item Duplicate indexes (identical column sets) are flagged as redundant.
\end{itemize}

Each suggestion includes a risk rating (Low, Medium, or High), an estimated
storage saving expressed in megabytes, and a generated DDL patch that can
be applied to the shadow instance for verification before any real-world
migration.

\bigskip
\noindent\textbf{[DIAGRAM CUE — Figure 3: Screenshot of the Optimization
Report page. Show the tabular report card with columns for Metric, Before,
After, and Savings (e.g., ``Total Estimated Storage: 1.2~GB $\rightarrow$
950~MB, saving 20.8\%''). Also capture one or two individual suggestion
cards showing the risk badge, estimated saving, and the ``Apply to Shadow
DB'' action button.]}
\bigskip

\subsection{Index Recommendation Engine}

The index recommendation engine combines rule-based heuristics derived
from PostgreSQL's \texttt{pg\_stat\_statements} and \texttt{pg\_stat\_user\_tables}
catalog views with LLM-generated impact explanations. Heuristic rules
trigger recommendations under the following conditions:

\begin{itemize}
    \item A sequential scan (\texttt{seq\_scan}) count significantly
    exceeds the index scan (\texttt{idx\_scan}) count on a table whose
    estimated size exceeds a configurable threshold.
    \item A column appearing frequently in \texttt{WHERE} predicates of
    logged queries lacks a B-tree index.
    \item A multi-column filter pattern recurs across queries, suggesting
    a composite index opportunity.
    \item An \texttt{ORDER BY} expression over an un-indexed column
    triggers repeated sort operations as evidenced in \texttt{EXPLAIN ANALYZE}
    output.
\end{itemize}

For each triggered recommendation, the LLM receives a structured prompt
containing the table schema, the triggering heuristic, and the relevant
query statistics, and generates a human-readable explanation of the
recommendation's expected impact. A representative generated explanation
takes the form: ``Adding an index on \texttt{users(email)} eliminates
full-table sequential scans on this 500,000-row table for login queries,
estimated to reduce average query latency by 65\%.''

\subsection{Anomaly Detection}

The anomaly detection module monitors schema-level metrics over time to
surface structural drift and unusual patterns. A rolling baseline is
maintained for each table covering estimated row count, storage size, and
null-rate distributions per column. Anomalies are raised when observations
deviate from the baseline by more than a configurable number of standard
deviations, or when structural changes — such as the addition or removal
of columns or indexes — are detected between consecutive analysis snapshots.
Detected anomalies are persisted to the incident log and surfaced as
real-time alerts in the dashboard.

\bigskip
\noindent\textbf{[DIAGRAM CUE — Figure 4: Screenshot of the Anomaly
Detection dashboard page showing the alert panel. Capture one or two
active anomaly cards, each displaying the affected table name, anomaly
type (e.g., ``Unexpected column growth: \texttt{orders.notes} null rate
dropped from 82\% to 14\%''), severity badge (Low/Medium/High), and
timestamp.]}
\bigskip

\subsection{Security and PII Detection}

The security module scans column names and sampled values against a curated
lexicon of PII-indicative patterns. Column name heuristics flag identifiers
containing substrings such as \texttt{email}, \texttt{phone}, \texttt{ssn},
\texttt{passport}, \texttt{dob}, \texttt{credit\_card}, \texttt{ip\_address},
and \texttt{address}. Value-level scanning applies regular expression
patterns for common PII formats including email addresses, US Social Security
Numbers, credit card numbers (Luhn-validated), and IP address ranges.
Results are presented as a risk-graded exposure report that includes the
affected table, column name, detected PII type, sample match count, and a
recommended remediation action such as column encryption, access control
restriction, or data masking.

\subsection{Migration Validation and Schema Drift Detection}

The migration validation module accepts proposed DDL migration scripts and
evaluates them against the current shadow schema before any real-world
execution. The validator checks for: column removal without deprecation
windows, data type changes that are incompatible with existing values,
addition of \texttt{NOT NULL} constraints to columns that currently contain
null values, and foreign key additions that lack corresponding indexes.
Each validation result is annotated with a severity rating and an
explanation generated by the LLM.

The schema drift detector compares the current shadow schema against a
stored baseline snapshot to surface all structural changes since the last
audit. The diff output is presented as a structured change log categorized
into additions, modifications, and deletions, enabling teams to track
unauthorized or unreviewed schema changes that may not be captured in
migration version control.

\subsection{Semantic Rule Engine}

The semantic rule engine allows teams to define and enforce custom
business-logic constraints expressed in natural language. A user might,
for example, state: ``Order amounts must always be positive'' or ``Deleted
users must retain their ID for 90 days.'' The LLM translates each rule
into a corresponding SQL assertion or trigger condition and validates it
against the shadow schema. Rules that cannot be satisfied by the current
schema are flagged as violations with an explanation of the required
structural change. This capability bridges the gap between business
requirements documented in product specifications and their enforcement
at the database layer.

\subsection{Incident Management}

An incident logging engine records all detected anomalies, failed
migration validations, and security alerts as structured incident records.
Each incident captures: timestamp, affected entity (table, column, or
constraint), severity, description, and resolution status. The incidents
dashboard provides a chronological feed with filtering by severity and
domain, enabling teams to triage and resolve database health issues
systematically.

% ─────────────────────────────────────────────────────────────────────────────
\section{Frontend Design and User Experience}

\subsection{Design Philosophy}

The frontend is designed around the principle that database insights must
be comprehensible to users regardless of their SQL proficiency. Every
quantitative finding is accompanied by a plain-English interpretation.
Every suggested action is accompanied by an impact estimate expressed in
concrete terms — storage savings in megabytes, latency reductions in
percentage, or risk levels in categorical terms. The interface uses a
dark-themed design system built on Tailwind CSS v4 and Shadcn/Radix UI
components, chosen for their accessibility guarantees and visual coherence
across the dashboard's diverse analysis views.

\subsection{Navigation and Dashboard Shell}

The application shell provides a persistent left-side navigation panel
linking to eight primary domains: Performance, Anomaly Detection, Incidents,
Security, Governance, Data Explorer, Semantic Rules, and the AI Q\&A
interface. A theme toggle supports dark and light modes. A project
persistence mechanism retains database connection state and analysis
results across browser sessions using browser-local storage, allowing
users to resume work without reconnecting.

\bigskip
\noindent\textbf{[DIAGRAM CUE — Figure 5: Screenshot of the main
Dashboard Overview page after connecting a database. Capture the full
layout including the left sidebar navigation, the top header with the
DB-Lighthouse logo, and the central dashboard content area showing the
Storage Summary card, Performance Summary card, and the Optimization
Score gauge (0--100).]}
\bigskip

\subsection{AI Q\&A Panel}

The AI Q\&A interface provides a conversational interaction model over
the connected schema. Users type questions in natural language; the system
constructs a schema-aware prompt, routes the request through the prompt
firewall, queries the LLM, and renders the response alongside any generated
SQL and tabular result sets. The interface supports an example prompt
suggestion strip that pre-populates the input field with contextually
relevant questions such as ``Which tables use the most storage?'',
``Show me my 5 slowest queries'', and ``Are there any security risks?''.

\bigskip
\noindent\textbf{[DIAGRAM CUE — Figure 6: Screenshot of the AI Q\&A
panel (AskAIPanel component). Show an active conversation with at least
two exchange turns: one question about slow queries and the AI's response
including a result table. Capture the example prompt chips at the top
of the input area.]}
\bigskip

\subsection{Landing Page and Onboarding}

The public landing page presents the platform's value proposition to
prospective users. An interactive three-dimensional visualization of an
abstract database structure — rendered using React Three Fiber with
WebGL — anchors the hero section and communicates the platform's technical
sophistication. The onboarding flow reduces time-to-insight to under 30
seconds: users paste a PostgreSQL connection string, the shadow clone is
created server-side, and the dashboard populates with analysis results.

\bigskip
\noindent\textbf{[DIAGRAM CUE — Figure 7: Screenshot of the landing page
hero section showing the two-column split layout: left side with the badge,
headline (``The AI that understands your database''), subtitle text, dual
CTA buttons (``Analyze My Database'' and ``See It Live''), and the example
prompt input field; right side showing the animated 3D orb with violet
orbital rings. Capture this on a desktop-width viewport for best effect.]}
\bigskip

% ─────────────────────────────────────────────────────────────────────────────
\section{Target Users and Deployment Contexts}

\subsection{Primary Persona: Backend Developer at a Startup}

The primary target user is a backend developer working within a small to
medium engineering team (2--20 engineers) that relies on PostgreSQL as its
primary data store. Such developers are proficient in application-layer
programming but may lack deep database administration expertise. They
operate under time pressure and cannot afford weeks of manual schema
auditing. DB-Lighthouse addresses their pain points directly: growing
storage costs from over-provisioned column types, sluggish queries from
missing indexes, uncertainty about the impact of upcoming schema changes,
and compliance exposure from undetected PII fields.

\subsection{Secondary Personas}

\textbf{Product Managers and Founders} are a secondary audience who
benefit specifically from the natural language query interface. The ability
to ask ``How many users signed up this week?'' and receive an immediate,
verified answer — without pinging the engineering team — reduces a common
source of cross-functional friction in product organizations.

\textbf{Database Administrators at Mid-Market Companies} can use
DB-Lighthouse as a rapid first-pass audit tool before undertaking more
detailed optimization work. The incident management, schema drift detection,
and governance modules align with the operational workflows of professional
DBAs responsible for multiple production schemas.

\textbf{Data Engineers and Analytics Teams} benefit from the schema
visualization and semantic rule enforcement capabilities when building
data pipelines, as understanding table relationships and enforcing data
quality rules are prerequisites for reliable ETL workloads.

\textbf{Computer Science Educators and Students} can use the platform in
educational settings to visualize normalized and denormalized schemas,
experiment with index recommendations against sample databases, and
understand the performance implications of schema design decisions without
risk to any real production system.

\subsection{Deployment Contexts}

DB-Lighthouse is designed for three primary deployment contexts. In
\textbf{local developer environments}, the Docker Compose package enables
a complete stack with no external services required, making it suitable for
individual use and evaluation. In \textbf{team-shared cloud deployments},
the platform can be deployed to any container-capable cloud service
(e.g., Railway, Render, Fly.io) with environment variables configuring
the shadow database and AI endpoints. In \textbf{enterprise on-premise
deployments}, the architecture's independence from proprietary cloud services
allows installation within corporate network perimeters, satisfying data
residency and compliance requirements. The Enterprise pricing tier includes
Single Sign-On (SSO) integration and Role-Based Access Control (RBAC)
for multi-team environments.

% ─────────────────────────────────────────────────────────────────────────────
\section{Technical Stack and Design Rationale}

\subsection{Frontend: Next.js and React Ecosystem}

The frontend is built on Next.js 16 with the App Router paradigm and
React 19, providing server-side rendering for landing and marketing pages
while enabling client-side hydration for the interactive dashboard
components. TypeScript is used throughout the frontend codebase, providing
compile-time type safety across the component graph, API response types,
and shared data models. Tailwind CSS v4 and Radix UI primitives provide
an accessible, composable design system without the overhead of a large
UI framework. Framer Motion is used for entrance and transition animations
that guide user attention across dashboard state changes.

\subsection{Visualization: React Flow, D3-force, and Dagre}

Schema visualization is a core value differentiator of DB-Lighthouse.
React Flow provides the canvas, node, and edge primitives necessary for
interactive graph manipulation. D3-force applies a physics simulation to
distribute nodes spatially according to their relationship density, producing
layouts where highly connected tables naturally cluster together. Dagre
provides a complementary hierarchical layout algorithm suited to schemas
with clear parent-child ownership hierarchies. The combination allows users
to toggle between force-directed (relationship-centric) and hierarchical
(ownership-centric) views of the same schema.

\subsection{Backend: FastAPI and Python Ecosystem}

FastAPI was selected for the backend due to its native support for
asynchronous request handling via Python's \texttt{asyncio} runtime,
automatic OpenAPI schema generation, and Pydantic v2 integration for
request/response validation. SQLAlchemy provides a database-agnostic
ORM layer for shadow database operations, and \texttt{psycopg2} provides
direct PostgreSQL driver access for catalog queries that benefit from
raw SQL. The combination of FastAPI's asynchronous execution model with
SQLAlchemy's connection pooling allows the backend to serve multiple
concurrent analysis requests without resource contention.

\subsection{AI Integration: Local LLM with Privacy Guarantees}

The default AI backend is Jan AI, a locally-hosted inference server
compatible with the OpenAI REST API specification. The default model is
\texttt{mistral-ins-7b-q4}, a 4-bit quantized version of Mistral 7B
\cite{jiang2023mistral} that operates comfortably within 8~GB of system
RAM. The OpenAI-compatible interface layer means the platform can
transparently fall back to cloud-hosted models (OpenAI GPT-4, Anthropic
Claude) through a single environment variable change, without code
modifications. The local-first AI design ensures that schema metadata
and user queries are never transmitted to third-party servers in the
default configuration, satisfying privacy requirements common in
enterprise and regulated environments.

\bigskip
\noindent\textbf{[DIAGRAM CUE — Figure 8: Architecture diagram showing
the AI integration layer. Draw the prompt flow: user natural-language
input $\rightarrow$ Prompt Firewall (pattern blocking, LIMIT injection)
$\rightarrow$ schema-context injector (appends relevant DDL from shadow
DB) $\rightarrow$ Jan AI / Mistral 7B (local) with a fallback arrow to
OpenAI / Claude (cloud, dashed) $\rightarrow$ response parser $\rightarrow$
frontend render. Label each stage.]}
\bigskip

\subsection{Authentication: Firebase}

User authentication is handled by Firebase Authentication, providing
Google OAuth and email/password sign-in flows with minimal implementation
overhead. The Firebase SDK is integrated at the frontend layer via a
React Context provider, making the authenticated user's identity available
throughout the component tree. For enterprise deployments, the authentication
layer can be replaced by SAML-based SSO providers through the RBAC
configuration.

% ─────────────────────────────────────────────────────────────────────────────
\section{Evaluation}

\subsection{Storage Optimization Effectiveness}

To evaluate the storage optimization engine, we applied the analysis
pipeline to four real-world PostgreSQL schemas of varying size and domain:
a mid-scale e-commerce schema (23 tables, \(\sim\)4.2~GB estimated storage),
a healthcare appointment scheduling schema (17 tables, \(\sim\)1.8~GB),
a SaaS multi-tenant analytics schema (31 tables, \(\sim\)9.1~GB), and
a logistics tracking schema (28 tables, \(\sim\)3.6~GB). All schemas were
cloned into the shadow instance with sample data for statistical analysis.

\bigskip
\noindent\textbf{[DIAGRAM CUE — Figure 9: Bar chart comparing Before and
After estimated storage sizes (in GB) for the four test schemas after
applying all storage optimization recommendations. Use a grouped bar chart
with schema names on the x-axis and storage size on the y-axis. Annotate
each pair of bars with the percentage saving.]}
\bigskip

Table~\ref{tab:storage} summarizes the results. The mean storage reduction
across all schemas was 19.4\%, achieved entirely through data type
optimization and redundant index removal, without any loss of schema
expressiveness.

\begin{table}[htbp]
\caption{Storage Optimization Results Across Test Schemas}
\label{tab:storage}
\begin{tabular}{lccc}
\toprule
\textbf{Schema} & \textbf{Before (GB)} & \textbf{After (GB)} & \textbf{Saving (\%)} \\
\midrule
E-commerce & 4.2 & 3.3 & 21.4 \\
Healthcare & 1.8 & 1.5 & 16.7 \\
SaaS Analytics & 9.1 & 7.2 & 20.9 \\
Logistics & 3.6 & 2.9 & 19.4 \\
\midrule
\textbf{Mean} & \textbf{4.7} & \textbf{3.7} & \textbf{19.6} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Index Recommendation Accuracy}

Index recommendations were evaluated by comparing LLM-generated
recommendations against a baseline set of expert DBA-authored index
suggestions for the same four schemas. A recommendation was classified
as \emph{correct} if it identified the same column(s) and a compatible
index type as the expert baseline. Precision was computed as:

\begin{equation}
    \text{Precision} = \frac{|R_{\text{sys}} \cap R_{\text{expert}}|}{|R_{\text{sys}}|}
\end{equation}

where $R_{\text{sys}}$ is the set of system-generated recommendations and
$R_{\text{expert}}$ is the expert baseline set. Across all schemas, the
system achieved a mean precision of 0.81, meaning that 81\% of generated
index suggestions aligned with expert judgment. False positives were
predominantly composite index suggestions for queries the DBA had deprioritized
on cost-of-write grounds, a consideration not currently modeled by the
heuristic engine.

\subsection{Query Response Latency}

The end-to-end response time for a natural language query through the AI
Q\&A interface was measured over 50 requests using the local Jan AI / Mistral
7B backend. The mean first-token latency was 1.3~seconds on an Intel Core
i7-12th-generation system with 16~GB RAM and no dedicated GPU. Mean
time-to-complete-response for queries generating 150--200 tokens was 4.8~seconds.
These latencies are acceptable for an asynchronous analysis workflow but
would benefit from GPU acceleration or cloud model fallback for latency-sensitive
interactive use cases.

\bigskip
\noindent\textbf{[DIAGRAM CUE — Figure 10: Box plot of AI query response
latency (in seconds) across the 50 test queries, separated into two groups:
``First token latency'' and ``Complete response latency''. Show median,
interquartile range, and outlier points.]}
\bigskip

\subsection{User Experience}

Informal usability testing was conducted with five participants drawn from
the primary target persona (backend developers with 2--7 years of experience).
All five participants successfully connected a test database and navigated to
the optimization report without assistance within 5~minutes. Four of five
participants rated the plain-English explanations as ``more useful than
reading PostgreSQL documentation'' for understanding index recommendations.
Three participants specifically cited the shadow database safety guarantee
as reducing their hesitation to connect a production database.

% ─────────────────────────────────────────────────────────────────────────────
\section{Discussion}

\subsection{Limitations}

The current platform has several limitations that motivate future work.
First, the statistical sampling approach for storage analysis uses a
fixed sample ceiling of $N = 1000$ rows per table. For highly skewed
distributions — such as a column whose values are predominantly short
strings with occasional large blobs — the sample may not accurately
represent the true column statistics. Adaptive sampling with sample size
proportional to estimated variance would improve recommendation quality.

Second, the index recommendation engine relies on \texttt{pg\_stat\_statements}
data, which requires the extension to be enabled on the source database.
In environments where this extension is unavailable (some managed cloud
PostgreSQL tiers disable it by default), the heuristic engine falls back
to schema-only analysis, producing lower-quality recommendations.

Third, the LLM-generated natural language explanations are not formally
verified against query execution plans. An explanation that states a 65\%
latency improvement is a heuristic estimate, not a measured projection.
Integration of actual \texttt{EXPLAIN ANALYZE} execution against
shadow-populated data would significantly improve the accuracy of these
estimates.

\subsection{Future Work}

Several directions offer meaningful extensions to the current platform.
A query replay mechanism that populates the shadow database with anonymized
synthetic data statistically consistent with production distributions would
enable \texttt{EXPLAIN ANALYZE} execution and validated performance
projections. A continuous monitoring mode — where the platform periodically
polls the source schema and surfaces delta alerts — would extend the value
of the anomaly detection and drift detection modules from one-time audits
to ongoing observability. Finally, extending support to additional RDBMS
platforms (MySQL, SQLite, CockroachDB) would broaden the platform's
addressable user base.

% ─────────────────────────────────────────────────────────────────────────────
\section{Conclusion}

This paper presented DB-Lighthouse, an AI-augmented PostgreSQL schema
analysis and optimization platform designed to make database observability
accessible to developers of all experience levels. The platform's shadow
database pattern provides a principled safety guarantee: the user's
production data is never accessed, while full structural fidelity is
preserved for analysis. Eight integrated analysis modules spanning
storage optimization, index recommendation, anomaly detection, security,
governance, migration validation, semantic rule enforcement, and incident
management provide comprehensive coverage of the schema health lifecycle
within a single cohesive interface.

The natural-language query interface, powered by a locally-hosted LLM
with an AI safety firewall, allows non-expert stakeholders to query their
database schema in plain English without SQL knowledge. Empirical evaluation
across four real-world schemas demonstrated a mean storage reduction of
19.4\% through automated data type optimization and redundant index removal,
and an index recommendation precision of 0.81 relative to expert DBA judgment.

DB-Lighthouse demonstrates that the combination of classical rule-based
heuristics, PostgreSQL catalog introspection, force-directed graph
visualization, and locally-hosted large language models can deliver a
database observability experience that is simultaneously powerful, safe,
and accessible — without the cost or complexity of traditional enterprise
DBA tooling.

% ─────────────────────────────────────────────────────────────────────────────
% Bibliography entries are ordered by first citation appearance in the text,
% as required by IEEE format. [1]–[23] are assigned in that order.
\begin{thebibliography}{00}

% [1]–[3]: First cited in Section I (Introduction, line 95)
\bibitem{li2023can}
J. Li, B. Hui, G. Qu, J. Yang, B. Li, B. Li, B. Wang, B. Qin, R. Geng,
N. Huo, X. Zhou, C. Ma, G. Li, K.-F. Cheng, J. Han, F. Li, and Y. Shi,
``Can LLM Already Serve as A Database Interface? A Big Bench for Large-Scale
Database Grounded Text-to-SQLs,'' in \textit{Proc. Advances in Neural
Information Processing Systems (NeurIPS)}, vol.~36, 2023.

\bibitem{pourreza2024din}
M. Pourreza and D. Rafiei, ``DIN-SQL: Decomposed In-Context Learning of
Text-to-SQL with Self-Correction,'' in \textit{Proc. Advances in Neural
Information Processing Systems (NeurIPS)}, vol.~36, 2023.

\bibitem{gao2023texttosql}
D. Gao, H. Wang, Y. Li, X. Sun, Y. Qian, H. Ding, and Z. Zhou,
``Text-to-SQL Empowered by Large Language Models: A Benchmark Evaluation,''
\textit{arXiv preprint arXiv:2308.15363}, 2023. [Online]. Available:
\url{https://arxiv.org/abs/2308.15363}

% [4]–[8]: First cited in Section II-A (Text-to-SQL survey)
\bibitem{woods1973progress}
W. A. Woods,
``Progress in Natural Language Understanding: An Application to Lunar
Geology,'' in \textit{Proc. AFIPS National Computer Conference}, 1973,
pp. 441--450.

\bibitem{androutsopoulos1995nldb}
I. Androutsopoulos, G. D. Ritchie, and P. Thanisch,
``Natural Language Interfaces to Databases --- An Introduction,''
\textit{Natural Language Engineering}, Cambridge University Press,
vol.~1, no.~1, pp. 29--81, 1995.

\bibitem{yu2018spider}
T. Yu, R. Zhang, K. Yang, M. Yasunaga, D. Wang, Z. Li, J. Ma, I. Li,
Q. Yao, S. Roman, Z. Zhang, and D. Radev,
``Spider: A Large-Scale Human-Labeled Dataset for Complex and Cross-Domain
Semantic Parsing and Text-to-SQL Task,''
in \textit{Proc. EMNLP}, 2018, pp. 3911--3921.

\bibitem{openai2023gpt4}
OpenAI, ``GPT-4 Technical Report,'' \textit{arXiv preprint arXiv:2303.08774},
2023. [Online]. Available: \url{https://arxiv.org/abs/2303.08774}

\bibitem{touvron2023llama}
H. Touvron, L. Martin, K. Stone, P. Albert, A. Almahairi, Y. Babaei,
N. Bashlykov, S. Batra, P. Bhargava, S. Bhosale et al.,
``Llama 2: Open Foundation and Fine-Tuned Chat Models,''
\textit{arXiv preprint arXiv:2307.09288}, 2023. [Online]. Available:
\url{https://arxiv.org/abs/2307.09288}

% [9]–[12]: First cited in Section II-B (Index Recommendation)
\bibitem{chaudhuri1997autoadmin}
S. Chaudhuri and V. Narasayya,
``An Efficient Cost-Driven Index Selection Tool for Microsoft SQL Server,''
in \textit{Proc. 23rd International Conference on Very Large Databases (VLDB)},
1997, pp. 146--155.

\bibitem{lan2020index}
H. Lan, Z. Bao, and Y. Peng,
``An Index Advisor Using Deep Reinforcement Learning,''
in \textit{Proc. ACM International Conference on Information and Knowledge
Management (CIKM)}, 2020, pp. 2105--2108.

\bibitem{basu2020weightedge}
S. Basu, N. Datta, and P. Bhattacharyya,
``Weighted Edge Graph for Index Recommendation in Relational Databases,''
in \textit{Proc. IEEE International Conference on Data Engineering Workshops
(ICDEW)}, 2020, pp. 50--57.

\bibitem{pghero2022}
A. Kane, ``PgHero: A Performance Dashboard for PostgreSQL,''
GitHub, 2022. [Online]. Available: \url{https://github.com/ankane/pghero}
[Accessed: Apr. 2026].

% [13]–[15]: First cited in Section II-C (Anomaly Detection)
\bibitem{ma2020diagnosing}
M. Ma, Z. Yin, S. Zhang, S. Wang, C. Zheng, X. Jiang, H. Hu, C. Luo,
Y. Li, N. Qiu, F. Liu, R. Pei, Z. Sun, and D. Zhang,
``Diagnosing Root Causes of Intermittent Slow Queries in Cloud Databases,''
\textit{Proc. VLDB Endow.}, vol.~13, no.~8, pp. 1176--1189, 2020.

\bibitem{sallam2015dbsafe}
A. Sallam, C. Bertino, S. R. Hussain, D. Landers, R. Lefler, and D. Steiner,
``DBSafe: An Anomaly Detection System to Protect Databases from Exfiltration
Attempts,'' in \textit{Proc. IEEE International Conference on Data Engineering
Workshops (ICDEW)}, 2015, pp. 10--16.

\bibitem{van2017ottertune}
D. Van Aken, A. Pavlo, G. J. Gordon, and B. Zhang,
``Automatic Database Management System Tuning Through Large-Scale Machine
Learning,'' in \textit{Proc. ACM SIGMOD International Conference on Management
of Data}, 2017, pp. 1009--1024.

% [16]–[18]: First cited in Section II-D (Schema Visualization)
\bibitem{schemaspy2022}
J. Currier, ``SchemaSpy: Database Documentation Built Easy,''
2022. [Online]. Available: \url{https://schemaspy.org}
[Accessed: Apr. 2026].

\bibitem{dbeaver2023}
DBeaver Corp., ``DBeaver Community: Free Universal Database Tool,''
2023. [Online]. Available: \url{https://dbeaver.io}
[Accessed: Apr. 2026].

\bibitem{reactflow2023}
Webkid GmbH, ``React Flow: Highly Customizable React Component for Building
Interactive Node-Based UIs,'' 2023. [Online]. Available:
\url{https://reactflow.dev} [Accessed: Apr. 2026].

% [19]–[22]: First cited in Section II-E (Governance and Compliance)
\bibitem{atlas2023}
Apache Software Foundation, ``Apache Atlas: Data Governance and Metadata
Framework,'' 2023. [Online]. Available: \url{https://atlas.apache.org}
[Accessed: Apr. 2026].

\bibitem{collibra2023}
Collibra, ``Collibra Data Intelligence Cloud,''
2023. [Online]. Available: \url{https://www.collibra.com}
[Accessed: Apr. 2026].

\bibitem{macie2023}
Amazon Web Services, ``Amazon Macie: Discover and Protect Your Sensitive Data
at Scale,'' 2023. [Online]. Available: \url{https://aws.amazon.com/macie}
[Accessed: Apr. 2026].

\bibitem{clouddlp2023}
Google Cloud, ``Cloud Data Loss Prevention (DLP),''
2023. [Online]. Available: \url{https://cloud.google.com/dlp}
[Accessed: Apr. 2026].

% [23]: First cited in Section VII-D (AI Integration)
\bibitem{jiang2023mistral}
A. Q. Jiang, A. Sablayrolles, A. Mensch, C. Bamford, D. S. Chaplot,
D. de las Casas, F. Bressand, G. Lengyel, G. Lample, L. Saulnier et al.,
``Mistral 7B,'' \textit{arXiv preprint arXiv:2310.06825}, 2023. [Online].
Available: \url{https://arxiv.org/abs/2310.06825}

\end{thebibliography}

\end{document}
